---
title: "Observed vs Logged Event Support"
author: "Robert Reid"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(IRanges)
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This document evaluates how well observed recombination events from a simulation
run are supported by the underlying logged true events. Three event types are
assessed: **NCO-GC**, **CO-GC**, and **CO-terminal**.

Support criteria differ by event type:

- **NCO-GC**: supported if a logged NCO event in the same cell shares an exact
  start and end position (`semi_join` on cell, start, end).
- **CO-GC**: supported if a logged CO event in the same cell is an exact range
  match (`pcompare` code = 0).
- **CO-terminal**: supported if a logged CO event in the same cell either shares
  a boundary consistent with the side of the centromere (codes -1, -3, 1, 3) or
  is directly adjacent with no gap (codes -5, 5). Directionality is enforced:
  events left of the centromere must share a common end or be adjacent on the
  right; events right of the centromere must share a common start or be adjacent
  on the left.


__Need to add DCO support__

---

## Configuration

Edit this section to point to your input files and centromere positions.

```{r config}
# Single Input files
#obs_file <- "test_output/obs2.tsv"
#log_file <- "test_output/log2.tsv"

# Folder of files

obs_files = list.files("batch_ns_1/",pattern = "^obs",full.names = TRUE)
log_files = list.files("batch_ns_1/",pattern = "^log",full.names = TRUE)

# Centromere positions: a named numeric vector where names are chromosome IDs
# and values are the centromere midpoint position on that chromosome.
# Replace with your actual centromere positions.
cen_positions <- c(
  "1"  = 152253,
  "2"  = 239200,
  "3"  = 112830,
  "4"  = 450091,
  "5"  = 151545,
  "6"  = 149655,
  "7"  = 497413,
  "8"  = 106304,
  "9"  = 356259,
  "10" = 436785,
  "11" = 439601,
  "12" = 151404,
  "13" = 274273,
  "14" = 629575,
  "15" = 326803,
  "16" = 556555
  # add chromosomes as needed
)
```

---

## Load Data

```{r load}
#obs <- read_delim(obs_file, delim = "\t", show_col_types = FALSE) |>
#  separate_wider_delim(event, delim = "-", names = c("co", "event_type"),
#                       too_many = "merge")
#
#log_events <- read_delim(log_file, delim = "\t", show_col_types = FALSE)

# load folder of files
expected_cells = 200 # this is should be set a bit above the max number of cells you
                     #. expect in a data file

obs = map(obs_files,\(f) read_delim(f, delim = "\t", show_col_types = FALSE)) |>
  imap(\(df, i) mutate(df, file_idx = i)) |>
  bind_rows() |>
  dplyr::rename(ocn = cell) |> # ocn = old cell number
  group_by(file_idx) |>
  mutate(cell = (cur_group_id() -1 ) * expected_cells + ocn) |>
  ungroup() |>
  separate_wider_delim(event, delim = "-", names = c("co", "event_type"),
                       too_many = "merge")

log_events = map(log_files,read_delim, delim = "\t", show_col_types = FALSE) |>
  imap(\(df, i) mutate(df, file_idx = i)) |>
  bind_rows() |>
  dplyr::rename(ocn = cell) |> # ocn = old cell number
  group_by(file_idx) |>
  mutate(cell = (cur_group_id() -1 ) * expected_cells + ocn) |>
  ungroup()

```

---

## Helper Functions

```{r helpers}
# --- DCO left-margin valid codes ---
# The left margin of the LOH region is matched when a logged CO shares or is
# adjacent to the observed start on its right side (logged range left of observed).
dco_left_valid <- function(code, obs_start, obs_end, log_start, log_end) {
  code %in% c(1, 3, 5)
}

# --- DCO right-margin valid codes ---
# The right margin is matched when a logged CO shares or is adjacent to the
# observed end on its left side (logged range right of observed).
dco_right_valid <- function(code, obs_start, obs_end, log_start, log_end) {
  code %in% c(-1, -3, -5)
}

# --- DCO support ---
# An observed DCO is supported if there exists at least one logged CO supporting
# its left margin AND at least one logged CO supporting its right margin, and
# both of those logged COs share the same generation (gen).
#
# Because gen is only in the log file, we run check_co_support twice (once per
# margin) using a modified version that also returns the matching log row index
# so we can retrieve gen from logged_co.
check_dco_support <- function(obs_dco, log_events) {
  logged_co <- log_events |>
    filter(type == "CO") |>
    select(cell, start, end, gen) |>
    mutate(log_row = row_number())

  logged_co_ranges <- IRanges(
    start = logged_co$start,
    end   = logged_co$end,
    names = as.character(logged_co$cell)
  )

  obs_ranges <- IRanges(
    start = obs_dco$start,
    end   = obs_dco$end,
    names = as.character(obs_dco$cell)
  )

  common_cells <- intersect(names(obs_ranges), names(logged_co_ranges))

  if (length(common_cells) == 0) {
    return(obs_dco |> mutate(n_supporting = 0L))
  }

  # For each cell, build a comparison table of all obs x log pairs with codes
  all_pairs <- map(common_cells, function(cell_id) {
    obs_sub <- obs_ranges[names(obs_ranges) == cell_id]
    log_sub <- logged_co_ranges[names(logged_co_ranges) == cell_id]

    # Retrieve gen for each log entry in this cell
    log_cell <- logged_co |> filter(as.character(cell) == cell_id)

    idx <- expand.grid(a = seq_along(obs_sub), b = seq_along(log_sub))
    cmp <- pcompare(obs_sub[idx$a], log_sub[idx$b])

    obs_row_idx <- which(as.character(obs_dco$cell) == cell_id)

    data.frame(
      obs_row   = obs_row_idx[idx$a],
      log_b     = idx$b,
      code      = cmp,
      gen       = log_cell$gen[idx$b],
      cell      = cell_id,
      stringsAsFactors = FALSE
    )
  }) |>
    bind_rows()

  # Split into left-margin and right-margin candidate pairs
  left_candidates <- all_pairs |>
    filter(dco_left_valid(code, NA, NA, NA, NA)) |>
    select(obs_row, gen_left = gen)

  right_candidates <- all_pairs |>
    filter(dco_right_valid(code, NA, NA, NA, NA)) |>
    select(obs_row, gen_right = gen)

  # A DCO is supported if any (left, right) candidate pair shares the same gen
  supported_obs <- inner_join(left_candidates, right_candidates, by = "obs_row") |>
    filter(gen_left == gen_right) |>
    distinct(obs_row) |>
    mutate(is_supported = TRUE)

  obs_dco |>
    mutate(obs_row = row_number()) |>
    left_join(supported_obs, by = "obs_row") |>
    mutate(
      n_supporting = if_else(is_supported, 1L, 0L, missing = 0L),
      is_supported = replace_na(is_supported, FALSE)
    ) |>
    select(-obs_row, -is_supported)
}


# --- NCO-GC support ---
# An observed NCO-GC is supported if a logged NCO event exists in the same cell
# with an identical start and end position.
# Returns n_supporting = 1 if matched, 0 otherwise (consistent with CO functions).
check_nco_support <- function(obs_nco, log_events) {
  logged_nco <- log_events |>
    filter(type == "NCO") |>
    select(cell, start, end)

  obs_nco |>
    mutate(
      n_supporting = map_int(seq_len(n()), function(i) {
        nrow(semi_join(
          obs_nco[i, c("cell", "start", "end")],
          logged_nco,
          by = c("cell", "start", "end")
        ))
      })
    )
}

# --- CO support via IRanges pcompare ---
# Returns a per-observation data frame with the count of supporting logged events.
check_co_support <- function(obs_co, log_events, valid_codes_fn) {
  logged_co <- log_events |>
    filter(type == "CO") |>
    select(cell, start, end)

  obs_ranges <- IRanges(
    start = obs_co$start,
    end   = obs_co$end,
    names = as.character(obs_co$cell)
  )

  log_ranges <- IRanges(
    start = logged_co$start,
    end   = logged_co$end,
    names = as.character(logged_co$cell)
  )

  common_cells <- intersect(names(obs_ranges), names(log_ranges))

  if (length(common_cells) == 0) {
    return(obs_co |> mutate(n_supporting = 0L))
  }

  # For each cell, compare every observed range against every logged range
  results <- map(common_cells, function(cell_id) {
    obs_sub <- obs_ranges[names(obs_ranges) == cell_id]
    log_sub <- log_ranges[names(log_ranges) == cell_id]

    idx <- expand.grid(a = seq_along(obs_sub), b = seq_along(log_sub))
    cmp <- pcompare(obs_sub[idx$a], log_sub[idx$b])

    # Retrieve the original obs row indices for this cell
    obs_row_idx <- which(as.character(obs_co$cell) == cell_id)

    data.frame(
      obs_row    = obs_row_idx[idx$a],
      log_idx    = idx$b,
      obs_start  = start(obs_sub)[idx$a],
      obs_end    = end(obs_sub)[idx$a],
      log_start  = start(log_sub)[idx$b],
      log_end    = end(log_sub)[idx$b],
      code       = cmp,
      cell       = cell_id,
      stringsAsFactors = FALSE
    )
  }) |>
    bind_rows()

  # Apply the caller-supplied filter to keep only valid comparisons
  supported_pairs <- results |>
    filter(valid_codes_fn(code, obs_start, obs_end, log_start, log_end))

  # Count supporting logged events per observed range
  support_counts <- supported_pairs |>
    count(obs_row, name = "n_supporting")

  obs_co |>
    mutate(obs_row = row_number()) |>
    left_join(support_counts, by = "obs_row") |>
    mutate(n_supporting = replace_na(n_supporting, 0L)) |>
    select(-obs_row)
}

# --- CO-GC valid codes ---
# Exact match only (pcompare code 0).
co_gc_valid <- function(code, obs_start, obs_end, log_start, log_end) {
  code == 0
}

# --- CO-terminal valid codes (with directionality) ---
# An observed terminal CO is left of the centromere if its midpoint is < CEN.
# Left of CEN  : the observed range should share a common END with the logged
#                range (code -1, -3) or be adjacent on its right (code -5).
# Right of CEN : the observed range should share a common START with the logged
#                range (code 1, 3) or be adjacent on its left (code 5).
make_co_terminal_valid <- function(cen_pos) {
  function(code, obs_start, obs_end, log_start, log_end) {
    obs_mid    <- (obs_start + obs_end) / 2
    left_of_cen <- obs_mid < cen_pos
    case_when(
      left_of_cen  & code %in% c(0,-1, -3, -5) ~ TRUE,
      !left_of_cen & code %in% c(0, 1,  3,  5) ~ TRUE,
      TRUE ~ FALSE
    )
  }
}
```

> **Note on `pcompare` codes for CO-terminal:**
> Negative codes mean the observed range precedes or extends left of the logged
> range; positive codes mean it extends right. Codes ±1 and ±3 indicate a shared
> boundary; codes ±5 indicate direct adjacency with no gap.

---

## Compute Support

```{r compute}
# --- DCO ---
dco_results <- obs |>
  filter(co == "DCO") |>
  select(cell, chrom, start, end) |>
  check_dco_support(log_events)

# --- NCO-GC ---
nco_results <- obs |>
  filter(co == "NCO", event_type == "GC") |>
  select(cell, chrom, start, end) |>
  check_nco_support(log_events)

# --- CO-GC ---
co_gc_results <- obs |>
  filter(co == "CO", event_type == "GC") |>
  select(cell, chrom, start, end) |>
  check_co_support(log_events, valid_codes_fn = co_gc_valid)

# --- CO-terminal ---
# Groups by chromosome so the correct centromere position is applied to each.
# chrom_num is a working column used only for the cen_positions lookup and is
# dropped before binding.
co_terminal_results <- obs |>
  filter(co == "CO", event_type == "terminal") |>
  select(cell, chrom, start, end) |>
  mutate(
    chrom_num = as.numeric(as.roman(chrom)),
    cen_pos   = cen_positions[as.character(chrom_num)]
  ) |>
  group_by(chrom) |>
  group_modify(~ {
    cen <- unique(.x$cen_pos)
    check_co_support(
      obs_co         = select(.x, -cen_pos, -chrom_num),
      log_events     = log_events,
      valid_codes_fn = make_co_terminal_valid(cen)
    )
  }) |>
  ungroup()
```

---

## Per-Observation Detail

```{r detail}
# All three result data frames now share the same columns:
# cell, chrom, start, end, n_supporting
# chrom is carried through so per-chromosome breakdowns are possible downstream.
detail <- bind_rows(
  nco_results         |> mutate(event_type = "NCO-GC"),
  co_gc_results       |> mutate(event_type = "CO-GC"),
  co_terminal_results |> mutate(event_type = "CO-terminal"),
  dco_results         |> mutate(event_type = "DCO")
) |>
  mutate(is_supported = n_supporting > 0)

detail
```

---

## Summary Table

```{r summary}
# Rolled-up summary by event type
summary_tbl <- detail |>
  group_by(event_type) |>
  summarise(
    n_observed    = n(),
    n_supported   = sum(is_supported),
    pct_supported = round(100 * n_supported / n_observed, 1),
    .groups = "drop"
  )

#summary_tbl

# Per-chromosome breakdown (useful once multi-chromosome simulations are in use)
summary_by_chrom <- detail |>
  group_by(chrom, event_type) |>
  summarise(
    n_observed    = n(),
    n_supported   = sum(is_supported),
    pct_supported = round(100 * n_supported / n_observed, 1),
    .groups = "drop"
  )

summary_by_chrom

failed <- detail |>
  filter(!is_supported) |>
  select(event_type, cell, chrom, start, end, n_supporting) |>
  arrange(event_type, chrom, cell, start)

failed
```
