---
title: "Recombination Modeling"
author: "Robert Reid"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
theme_set(theme_bw())
```

## WT cells

Recombination modeling of wild type cells based on the work of Sui
and Petes, 2020.
Recombination was simulated with functions in `simRec.py` and
multiple cells were simulated using `simRec_batch.py`.
The lower limit of basal recombination rate for unperturbed growth
from Sui and Petes of 3.7e-10 per generation per basepair was
used for simulation of 88 cells at 3000 generations per batch.
100 simulations were performed.
Accumulated data will be loaded and plotted.

#### Load data

```{r load data, echo=FALSE, message=FALSE, warning=FALSE}
#log_fn = list.files("batch/",pattern = "log\\d{2}.tsv",full.names = TRUE)
obs_fn = list.files("batch10/",pattern = "obs\\d{2}.tsv",full.names = TRUE)

#tf = read_delim(f[1]) |>
#  mutate (
#    class = factor(event,levels = c("CO-GC","CO-terminal","NCO-GC"))
#  )
#
#sf = tf |> count(class, .drop = FALSE)

bd = map(obs_fn,function(fn){
  fc = read_delim(fn) |>
    mutate(class = factor(event,levels = c("CO-GC","CO-terminal","NCO-GC","DCO-internal")))
  st = fc |> count(class, .drop = FALSE)
  return(st)
}) |>
  bind_rows()

```

Plot

```{r}
bd |>
  ggplot(aes(x=class,y=n)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.15)) +
  ylab("Number of events") +
  ylim(0,NA) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 15),
    axis.text = element_text(size = 12))

```

Summary

```{r}
bd |>
  group_by(class) |>
  summarise(
    median = median(n,na.rm = TRUE)
  )

bd |>
  group_by(class) |>
  summarise(
    max = max(n,na.rm = TRUE)
  )
```

