---
title: "Recombination Modeling"
author: "Robert Reid"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
theme_set(theme_bw())
```

## WT cells

Recombination modeling of wild type cells based on the work of Sui
and Petes, 2020.
Recombination was simulated with functions in `simRec.py` and
multiple cells were simulated using `simRec_batch.py`.
The lower limit of basal recombination rate for unperturbed growth
from Sui and Petes of 3.7e-10 per generation per basepair was
used for simulation of 88 cells at 3000 generations per batch.
100 simulations were performed.
Accumulated data will be loaded and plotted.

#### Load data

```{r load data, echo=FALSE, message=FALSE, warning=FALSE}
f = list.files("test_batch_5e10/",pattern = "*.tsv",full.names = TRUE)
bd = map(f,function(f){
  fc = read_delim(f)
  fct = table(fc$event)
  return(fct)
}) |>
  bind_rows() |>
  mutate(ILOH = `GC-CO` + `GC-NCO`) |>
  rename(TLOH = `CO-terminal`)

bdl = bd |>
  pivot_longer(cols = everything(),names_to = "event",values_to = "count")

```

Plot

```{r}
bdl |>
  filter(event %in% c("ILOH","TLOH")) |>
  ggplot(aes(x=event,y=count)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.15))
```

Summary

```{r}
bdl |>
  group_by(event) |>
  summarise(
    median = median(count,na.rm = TRUE)
  )
```

